---
title: "Timeseries Analysis"
format: html
editor: visual
editor_options: 
  chunk_output_type: console
---

## Time series data

- What is it? Data collected over time.
- To better understand time series data, we will look at water quality data collected from 2018-2024 at UNH's Coastal Marine Lab

## Load and tidy data

```{r}
# Load packages
library(tidyverse)

# Load data
# data<-read.csv("data/UNH_CML_2018_2024.csv")

# time is in "utc"
# temperature is in "C"

#data<-data[-1,] # removing the first row


# Adjusting the format of my variables and making new variables
#data <- data |> 
#  mutate(time =  as.POSIXct(time, format = "%Y-%m-%dT%H:%M:%SZ", tz="UTC")) |> # converting my time column to datetime format
#  mutate(temperature = as.numeric(temperature)) |> # convert temp to numeric value
#  filter(!temperature == -999) |> # filtering out all rows that have -999 in the temperature column
#  mutate(datetime = time) # making a new column called datetime which has all of the info from the time col
  
```

### Making even MORE new variables

- These will be helpful for some of the plots I will be crearting later on

```{r}
#data<-data |> 
#  mutate(date = as.Date(datetime), # make date data from my datetime col
#         year = as.factor(year(datetime)), # extracting only the year
#         month = as.factor(month(datetime)), # extracting the month
#         doy = yday(datetime)) # creating a day of the year variable
```

## Time series data is comprised of:

1)  trend - long-term progression of dat
2)  seasonality - repeating patterns
3)  noise - random irregular fluctuations

Knowing these different componenets helps us understand patterns of the past and essential for forecasting (i.e., predicting into the future)

# Plotting our data

```{r}

#ggplot(data = data, aes(x = datetime, y = temperature))+ # plot datetime on the x axis and temperature on the y
#  geom_point(size = 0.1) # make this a scatter plot

# Is there a *trend*?
#plot<-ggplot(data = data, aes(x = datetime, y = temperature))+ #save my plot as an object called "plot"
#  geom_point(size = 0.1)+
# geom_smooth(method = "lm") #add a trend line
##plot # call/display my "plot"

library(plotly)
# Plottly makes your plots interactive
#ggplotly(plot) # display my plot using ggplotly()


# Lets look at different types of *seasonality* in our data.
#data |> 
#  group_by(year, doy) |> # group by year and doy (day of the year)
#  ggplot(aes(x = doy, y = temperature, color = year))+ # plot doy on x axis, and temp on y, color by year
#  labs(x ="Day of year" , y = "Temperature")+ # adjust lables 
#  facet_wrap(~ year) # separate plots based on year

# Let's make a time column to look at patterns across a DAY

library(hms) 

# Looking at daily pattern
#year2022<-data |> # create a new object year2022
#  filter(year == 2022) |> # filter out year 2022 
#  mutate(datetime = lubridate::round_date(datetime, unit = "hour")) #|> # round our datetime to hours
#  mutate(eastern_datetime = with_tz(datetime, tz = "America/New_York")) |> # converting my datetime to eastern
#  mutate(eastern_time = as_hms(eastern_datetime)) # extract only time
  
# Plotting year 2022 data
#year2022 |> 
#  group_by(eastern_time) |> # group by time/hour
#  ggplot(aes(x = eastern_time, y = temperature)) +
#  geom_point(size = 0.1)

# A little hard to interpret, let's take some summary stats
#year2022 |> 
#  group_by(eastern_time) |> 
#  summarise(mean_temperature = mean(temperature)) |> 
#  ggplot(aes(x = eastern_time, y = mean_temperature))+
#  geom_point()+
#  geom_line()

```

## Visualize our data in different ways

```{r}

# Heatmaps are great for time series data!
#data |> 
#  group_by(year, month) |> # group by year and month
#  summarise(mean_monthly_temp = mean(temperature)) |> # calculating mean monthly temp
#  ggplot()+
#  geom_tile(aes(fill = mean_monthly_temp, x = as.character(month), y = year))+ # create a tile plot, squares will be filled based on the mean_monthly_temp, month will be on the x axis and year will be on the y axis
#  labs(x = "Month", y = "Year", fill = "Mean monthly temperature (C)")+
#  scale_x_discrete(labels = month.abb)


```
